Quantitative Auditing of AI Fairness with Differentially Private Synthetic Data

### 2024-12-10 paper/paper.tex
Added todos

@@

@@

@@

\begin{table*}[h]
\caption{Fairness measures.}
\label{tab:measures}
\begin{tabular}{llc}
\toprule
\textbf{Category} & \textbf{Fairness Measure} & \textbf{Definition} \\
\midrule
\multirow{4}{*}{Independence} & Disparate Impact & $\frac{P[\hat{Y} = 1 | S \neq 1]}{P[\hat{Y} = 1 | S = 1]} \geq 1 - \epsilon$ \\
& Demographic Parity & $\abs{P[\hat{Y} = 1 | S = 1] - P[\hat{Y} = 1 | S \neq 1]} \leq \epsilon$ \\
& Conditional Statistical Parity & $\abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon$ \\
& Mean Difference & $\abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon$ \\
\multirow{4}{*}{Separation} & \multirow{2}{*}{Equalized Odds} & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
& & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
& Equal Opportunity & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
& Predictive Equality & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
\multirow{4}{*}{Sufficiency} & \multirow{2}{*}{Conditional Use Accuracy Equality} & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
& & $\abs{P[Y = 0 | S = 1, \hat{Y} = 0] - P[Y = 0 | S \neq 1, \hat{Y} = 0]} \leq \epsilon$ \\
& Predictive Parity & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
& Equal Calibration & $\abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon$ \\
\multirow{3}{*}{N/A} & Accuracy Equality & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
& Positive Balance & $\abs{E[\hat{V} | Y = 1, S = 1] - E[\hat{V} | Y = 1, S \neq 1]} \leq \epsilon$ \\
& Negative Balance & $\abs{E[\hat{V} | Y = 0, S = 1] - E[\hat{V} | Y = 0, S \neq 1]} \leq \epsilon$ \\
\bottomrule
\end{tabular}
\end{table*}

@@

\begin{definition}[R\'enyi Differential Privacy (RDP)]\label{def:rdp}
Let $P_{\mathbf{X}}$ denote the probability distribution induced by the random vector $\mathbf{X}$. A randomized mechanism $M$ satisfies $(\alpha,\gamma)$-RDP for $\alpha \geq 1$ and $\gamma \geq 1$ if, for all databases $\db_1 \sim \db_2$, we have
\[
D_\alpha(P_{M(\db_1)} \Vert P_{M(\db_2)}) \leq \gamma
\]
where $D_\alpha(P_1 \Vert P_2)$ is the R\'enyi divergence\cite{van2014renyi,van2010renyi,li2016renyi} of order $\alpha$ between probability distributions $P_1,P_2$ over $x$
\[
D_\alpha(P_1 \Vert P_2) := \frac{1}{\alpha - 1} \text{log} \int {P_1(x)}^\alpha {P_2(x)}^{1-\alpha} \text{d}x
\]
\end{definition}

@@

@@

@@

@@

We hence developed an alternative heuristic for the general-purpose workflow. As mentioned in Section~\ref{sec:prelim}, the mutual information of two random variables is bounded by the pair's respective Shannon entropy. Using this property, we add additional edges with weights exceeding a fraction of the minimum of these upper bounds. As a rule of thumb, we have found setting the fraction to be $0.1$ to be effective.

% \begin{algorithm}
% \caption{something something}
% \label{alg:heuristic}
% \begin{algorithmic}[1]
% \STATE Initialize $x \gets 0$
% \STATE Initialize $y \gets 1$
% \FOR{$i \gets 1$ to $n$}
%     \IF{$i$ is even}
%         \STATE $x \gets x + i$
%     \ELSE
%         \STATE $y \gets y \times i$
%     \ENDIF
% \ENDFOR
% \RETURN $x + y$
% \end{algorithmic}
% \end{algorithm}

@@

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{adult_mst}
      \caption{Markov random field for marginals of the Adult dataset.}
      \label{fig:adult_mst}
      \Description{Markov random field for marginals of the Adult dataset.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{compas_mst}
      \caption{Markov random field for marginals of the COMPAS dataset.}
      \label{fig:compas_mst}
      \Description{Markov random field for marginals of the COMPAS dataset.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{diabetes_mst}
      \caption{Markov random field for marginals of the Diabetes dataset.}
      \label{fig:diabetes_mst}
      \Description{Markov random field for marginals of the COMPAS dataset.}
  \end{subfigure}
      \caption{Markov random field for marginals of the experimented datasets.}
      \label{fig:msts}
\end{figure}

@@

@@

@@

@@

>>>>

% TODO: add more to intro

% TODO: lowercase Diff

@@

% TODO: ai woulld be found to be unfair by 3rd party. example: compas

@@

% TODO: to mitigate use synthetic data. principals of this

@@

% TODO: remove unconsidered

% TODO: rename equalized odds 1,2

\begin{table*}[h]
\caption{Fairness measures.}
\label{tab:measures}
\begin{tabular}{llc}
\toprule
\textbf{Category} & \textbf{Fairness Measure} & \textbf{Definition} \\
\midrule
\multirow{1}{*}{Independence}
% & Disparate Impact & $\frac{P[\hat{Y} = 1 | S \neq 1]}{P[\hat{Y} = 1 | S = 1]} \geq 1 - \epsilon$ \\
& Demographic Parity & $\abs{P[\hat{Y} = 1 | S = 1] - P[\hat{Y} = 1 | S \neq 1]} \leq \epsilon$ \\
% & Conditional Statistical Parity & $\abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon$ \\
% & Mean Difference & $\abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon$ \\
\multirow{2}{*}{Separation}
& \multirow{2}{*}{Equalized Odds} & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
& & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
% & Equal Opportunity & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
% & Predictive Equality & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
\multirow{2}{*}{Sufficiency}
& \multirow{2}{*}{Conditional Use Accuracy Equality} & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
& & $\abs{P[Y = 0 | S = 1, \hat{Y} = 0] - P[Y = 0 | S \neq 1, \hat{Y} = 0]} \leq \epsilon$ \\
% & Predictive Parity & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
% & Equal Calibration & $\abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon$ \\
\multirow{1}{*}{N/A}
& Overall Accuracy Equality & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
% & Positive Balance & $\abs{E[\hat{V} | Y = 1, S = 1] - E[\hat{V} | Y = 1, S \neq 1]} \leq \epsilon$ \\
% & Negative Balance & $\abs{E[\hat{V} | Y = 0, S = 1] - E[\hat{V} | Y = 0, S \neq 1]} \leq \epsilon$ \\
\bottomrule
\end{tabular}
\end{table*}

@@

\begin{definition}[R\'enyi Differential Privacy (RDP)]\label{def:rdp}
Let $P_{\mathbf{X}}$ denote the probability distribution associated with the random vector $\mathbf{X}$. A randomized mechanism $M$ satisfies $(\alpha,\gamma)$-RDP for $\alpha \geq 1$ and $\gamma \geq 1$ if, for all databases $\db_1 \sim \db_2$, we have
\[
D_\alpha(P_{M(\db_1)} \Vert P_{M(\db_2)}) \leq \gamma
\]
where $D_\alpha(P_1 \Vert P_2)$ is the R\'enyi divergence\cite{van2014renyi,van2010renyi,li2016renyi} of order $\alpha$ between probability distributions $P_1,P_2$ over $x$
\[
D_\alpha(P_1 \Vert P_2) := \frac{1}{\alpha - 1} \text{log} \int {P_1(x)}^\alpha {P_2(x)}^{1-\alpha} \text{d}x
\]
\end{definition}

@@

% TODO: reiterate intro

% TODO: goal of this framework

@@

% TODO: write more about these paragraphs

@@

% TODO: add a overarching example for methodologies

@@

% We hence developed an alternative heuristic for the general-purpose workflow. As mentioned in Section~\ref{sec:prelim}, the mutual information of two random variables is bounded by the pair's respective Shannon entropy. Using this property, we add additional edges with weights exceeding a fraction of the minimum of these upper bounds. As a rule of thumb, we have found setting the fraction to be $0.1$ to be effective.

% TODO:
% for synth to good : depend on domain knowledge, tractibility
% synth algos use dependencies
% we DO NOT use this. we do vanilla
% experiment didnt developed heuristics:
% 1. lack of domain knowledge
% 2. different prediction target. cannot tailor a priori

@@

% TODO: move this to their respective subsection

% TODO: reiterate methodologies

@@

% \begin{figure}
%   \centering
%   \begin{subfigure}[b]{0.3\textwidth}
%       \centering
%       \includegraphics[width=\textwidth]{adult_mst}
%       \caption{Markov random field for marginals of the Adult dataset.}
%       \label{fig:adult_mst}
%       \Description{Markov random field for marginals of the Adult dataset.}
%   \end{subfigure}
%   \hfill
%   \begin{subfigure}[b]{0.3\textwidth}
%       \centering
%       \includegraphics[width=\textwidth]{compas_mst}
%       \caption{Markov random field for marginals of the COMPAS dataset.}
%       \label{fig:compas_mst}
%       \Description{Markov random field for marginals of the COMPAS dataset.}
%   \end{subfigure}
%   \hfill
%   \begin{subfigure}[b]{0.3\textwidth}
%       \centering
%       \includegraphics[width=\textwidth]{diabetes_mst}
%       \caption{Markov random field for marginals of the Diabetes dataset.}
%       \label{fig:diabetes_mst}
%       \Description{Markov random field for marginals of the COMPAS dataset.}
%   \end{subfigure}
%       \caption{Markov random field for marginals of the experimented datasets.}
%       \label{fig:msts}
% \end{figure}

@@

% TODO: dont mention shape of graph

@@

% TODO:
first list formulas

count vars

theyre not part of marginals

they cant be computed

we can say less about fairness with these with less info

auditor need to know theyre aproxing scores

auditor need to keep in mind that these scores are for datasets and we preserve these scores thats it. scores may not be accurate. it doesnt mean its less easy to test fairness

@@

% TODO:
it works!

@@

% TODO: use = 1
